---
title: "Visualization Lab 3"
author: "Roshni Sundaramurthy (rossu809) & Brian Masinde (brima748)"
date: "30 September 2018"
output:
  html_document:
    df_print: paged
  html_notebook:
    theme: journal
  pdf_document: default
fontsize: 11pt
bibliography: references.bib
---

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)
library(seriation)
library(dplyr)
library(lattice)
library(plotly)
library(scales)
library(tidyr)
```

## Assignment 1

```{r echo=FALSE}
prices_earnings <- read.delim("prices-and-earnings.txt")
```

### 1.1

```{r subset}
# keep columns 1,2,5,6,7,9,10,16,17,18,19

p_e <- prices_earnings[, c(1,2,5,6,7,9,10,16,17,18,19)]

# set column 1 to row names.
rownames(p_e) <- p_e[[1]]
```


```{r scaling, echo = FALSE}
# Preliminary scaling step to question 
p_e_sc <- scale(p_e[,-1])

```

### 1.2

Without doing any reordering We cannot identify any clusters or outliers. The heatmap looks randomized.
```{r heat}
#p_e_sc %>% it was not possible to pipe the data since it is a matrix
  plot_ly(x =~colnames(p_e_sc), y =~rownames(p_e_sc),
    z = ~p_e_sc, type = "heatmap", 
    colors = colorRamp(c("black","red"))
  ) %>%
  layout(title =  "Heatmap of prices and earnings (Without ordering)",
         xaxis = list(title = "Price-Earnings Indicators", zeroline = FALSE),
         yaxis = list(title = "Cities", zeroline = FALSE)
  )
```

### 1.3

The ordering by euclidean distance produces a heat map that is easier to analyze. At first glance we can perceive four general regions of two groups. The first group heat map color tends towards higher price values (a brighter shade of red) while the second group tend towards lower price values (a darker shade of red/black). Although these groups can be seen in the correlation distance heat map, it is not as clear as the one produced by euclidean distance.

Based on the euclidean distance heat map, net wage tends to higher values from Dubai up towards Tokyo while the number of hours worked decrease. This is the opposite of cities like Delhi,Bankok and Seoul. Interestingly food costs are generally low in the cities with higher working hours. Caracas is an outlier because food costs are high while net wage and the number of hours worked remains low.

We used method "OLO" as the Hierarchical Clustering algorithm instead of "HC" method, because according to the documentation in seriation package and @[hahsler] the former does not optimize the Hamiltonian Path Length.

```{r euclidHC, echo = TRUE}
# seriation needs to permute rows and columns, thus distance by row and column
p_e_rdist <- dist(p_e_sc, method = "euclidean")

# by column
p_e_cdist <- dist(t(p_e_sc), method = "euclidean")

# make sure that results are reproducible
set.seed(1011)

# get orders of the row and col distances; Hamilton path length
ser_row_HC <- seriate(p_e_rdist, method = "OLO")

ser_col_HC <- seriate(p_e_cdist, method = "OLO")

order1 <- get_order(ser_row_HC)

order2 <-get_order(ser_col_HC )

# reorder
p_reord <- p_e_sc[rev(order1), order2]

# plot
plot_ly(x =~colnames(p_reord), y =~rownames(p_reord),
    z = ~p_reord, type = "heatmap", 
    colors = colorRamp(c("black","red"))
  ) %>%
  layout(title =  "Heatmap of prices and earnings (Euclid dist)",
         xaxis = list(title = "Price-Earnings Indicators", zeroline = FALSE),
         yaxis = list(title = "Cities", zeroline = FALSE)
  )
```

```{r euclidCOR, echo = FALSE}
# computing distance as one minus correlation

p_e_cor <- as.dist((1 - cor(p_e_sc)))

p_e_cor1 <- as.dist((1 - cor(t(p_e_sc))))

# set seed to ensure results are reproducible
set.seed(10111)

# get orders for columns and rows
ord1 <- get_order(seriate(p_e_cor, method = "OLO"))

ord2 <- get_order(seriate(p_e_cor1, method = "OLO"))

# reorder
p_reord2 <- p_e_sc[rev(ord2), ord1]

#plot
plot_ly(x =~colnames(p_reord2), y =~rownames(p_reord2),
    z = ~p_reord2, type = "heatmap", 
    colors = colorRamp(c("black","red"))
  ) %>%
  layout(title =  "Heatmap of prices and earnings (Cor dist)",
         xaxis = list(title = "Price-Earnings Indicators", zeroline = FALSE),
         yaxis = list(title = "Cities", zeroline = FALSE)
  )
```

### 1.4

From visual comparison of the heatmap produced by HC solver and TSP, we prefer the latter because there appears to be a separation along the anti diagonal such that cities that have similary high prices are placed along this diagonal. In the heatmap by the TSP solver this distinction is not quite clear.

```{r heatTSP}
# use p_e_rdist and p_e_cdist (euclidean distance)
# set seed
set.seed(11)

ser_row_TSP <- seriate(p_e_rdist, method = "TSP")

set.seed(111)

ser_col_TSP <- seriate(p_e_cdist, method = "TSP") 

# orders
ord_q4_1 <- get_order(ser_row_TSP)

order_q4_2 <-get_order(ser_col_TSP)

# reorder
p_reord_q4 <- p_e_sc[rev(ord_q4_1), order_q4_2]

# plot
plot_ly(x =~colnames(p_reord_q4), y =~rownames(p_reord_q4),
    z = ~p_reord_q4, type = "heatmap", 
    colors = colorRamp(c("black","red"))
  ) %>%
  layout(title =  "Heatmap of prices and earnings (Euclid dist- TSP)",
         xaxis = list(title = "Price-Earnings Indicators", zeroline = FALSE),
         yaxis = list(title = "Cities", zeroline = FALSE)
  )
```

TSP solver has shorter path length (119.09) compared to HC solver (121.967). Thus it does a better job of optimizing the Hamiltonian Path Length. For measure of effectiveness (ME) the TSP solver (ME = 653.2224) has advantage over the HC (ME = 652.4429). A higher ME implies a better arrangement of the dissimilarity matrix. For gradient measures, since objective is to increase the distance from main diagonal, we infer that HC (Gradient_raw = 65528) does a better job of achieving this compared to TSP solver(Gradient_raw = 39102).

```{r criterion, echo = TRUE}
# function creterion to compare unordered distance and ordered
# distance = p_e_rdist (row distance)
# use ser_row_HC because its a simulation, we need to compare the exact same results as used in heatmaps
# use ser_row_TSP, same reason as above 


result <- rbind(HC = criterion(p_e_rdist,ser_row_HC ), TSP  = criterion(p_e_rdist,or2))

result
```



## Assignment 2

```{r read data into R, echo=FALSE, message=FALSE}
df <- read.csv("adult.csv")
colnames(df) <- c("age","workclass","fnlwgt","edu","edu_num","marital_status","occupation","relationship","race","sex","cap_gain","cap_loss","hours_per_week","native_country","income_level")
#View(df)
```

### 2.1 Scatter and Trellis plots

The points in the scatter plot are very close to each other and so, it is difficult to interpret the results for both <=50K and >50K income levels.
The trellis plot shows both income levels in separate panels and so, it is quite easy to analyse the datas. The people with age range of 0-25 who earns <=50K are working for long hours than people earning >50K.

```{r Scatter and Trellis}
plot1 <- ggplot(df, aes(age,hours_per_week,color=income_level))+labs(title="Scatter plot",x="Age", y = "Hours per week")+
         geom_point(size=1)
plot1

plot2<-ggplot(df, aes(age,hours_per_week,color=income_level))+geom_point(size=1)+labs(title="Trellis plot",x="Age", y = "Hours per week")+
   facet_grid(income_level~.)
plot2
```

### 2.2 Density plot

The density curve is skewed to the right (right skewed distribution) for income level <=50K and thus not normally distributed.The median income level for <=50K is less than the mean. Whereas the the density curve for income of >50K is bimodal.
The trellis plot exhibits both income level for people with various marital status. The density curve for widower is normally distributed for both income levels. The people who were never married earnig <=50K lies within range of 17-40 age group.

```{r}
plot3 <- ggplot(df, aes(x=age, group=income_level, fill=income_level))+
         geom_density(alpha=0.5) + labs(title="Density plot of age")
plot3

plot4<-ggplot(df, aes(x=age,group=income_level,fill=income_level))+geom_density(alpha=0.5)+labs(title="Trellis plot og age")+facet_grid(marital_status~.)
plot4
```

### 2.3 3D-scatter plot and 2D-density plot

In 3D-scatter plot, the overlapping of data points around the same values makes it difficult to analyse the relationship between given three varaibles. In raster type 2D-density plot, every panels of age group 29-90 exhibits almost same outputs for capital loss ranging 1000-2000 except age group of 17-29.

```{r 3D scatter plot}
df1 <- filter(df, cap_loss!=0)
df1 %>% plot_ly(x = ~edu_num, y = ~age, z = ~cap_loss) %>%
  add_markers(colorscale = c('#BF382A', '#0C4B8E', '#1c1c1c'), showscale=TRUE) %>%
  layout(scene = list(xaxis = list(title = 'Education-num'),
                     yaxis = list(title = 'Age'),
                     zaxis = list(title = 'Capital loss')))
```

```{r 2D Density plot}
plot5 <- ggplot(df1, aes(x=edu_num,y=cap_loss))+stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE)+labs(title="2D raster type density plot")+theme(legend.position='none')+facet_grid(cut_number(df1$age,6)~.)
plot5
```

### 2.4 Trellis plots

The patterns in every panels of trellis plot looks almost similar except the last panel i.e, age group of 48-90. The capital loss seems to be quite high for both age group of 37-48 and 48-90.

```{r (a) using cut number}
plot6a<-ggplot(df, aes(x=edu_num,y=cap_loss,color=cap_loss))+geom_point()+labs(title="Trellis plot")+
  facet_grid(cut_number(df$age,4)~.)
plot6a
```


```{r (b) using shingles}
new_age = equal.count(df1$age, number=4,overlap=.1)
M <- matrix(unlist(levels(new_age)), ncol = 2, byrow = TRUE)
M1 <- data.frame(Lower=M[,1],Upper=M[,2], Interval=factor(1:nrow(M)))

index=c()
Class=c()
for(i in 1:nrow(M)){
  Cl=paste("[", M1$Lower[i], ",", M1$Upper[i], "]", sep="")
  ind=which(df$age>=M1$Lower[i] &df$age<=M1$Upper[i])
  index=c(index,ind)
  Class=c(Class, rep(Cl, length(ind)))
}

df2<-df[index,]
df2$Class<-as.factor(Class)

plot6b <- ggplot(df2, aes(x=edu_num,y=cap_loss,color=cap_loss))+ geom_point()+labs(title="Trellis plot (Shingles)")
  facet_grid(Class~.)
  plot6b
  
```
